Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface

### 2022-04-10 doc/AMS-15/AMS-15_paper/AMS_HGIS_els/AMS_HGIS_els.Rmd
Fix references, start on edits.

Radiocarbon Accelerator Mass Spectrometry (AMS) is a sensitive method for determining the age, provenance, and composition of carbon-containing materials. A primary disadvantage of the method is its cost, which often limits the number of samples analyzed. This forces researchers to carefully choose samples to be run and limits repeat measurements for error determination. One driver of the cost of AMS analysis is the time and effort required for sample preparation, including reduction of the sample to graphite. Analysis of CO~2~ directly via a gas-accepting AMS ion source eliminates the cost and effort of graphitization and returns results more rapidly. The NOSAMS USAMS system at the National Oceanographic AMS lab (NOSAMS) is based on a 3MV Tandetron accelerator [@Longworth2015] combined with a NEC 134-sample gas-capable MCSNICS source [@Xu2007] . We present here a system for semi-automated analysis of carbonate minerals by acid hydrolysis in vials and introduction to the ion source in a flow of helium via an open split.

@@

```{r fig1, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig1}Schematic of the open split inlet. Two valco 2-way valves provide the ability to admit helium and live or dead tank standards. The double needle is housed in a VICI cross and allows displacement helium to flow into the sample vial and the sample-helium mixture to flow to the open split, where a capillary delivers the mixture to the ion source.", echo = FALSE}
knitr::include_graphics("../figures/fig1_HGIS_open_split_drawing.pdf")
```

@@

Carbonate samples were weighed into 7 ml septa-sealed Vacutainer vials (Labco 768W). Sealed vials were evacuated, then 1 ml 85% phosphoric acid was added. The samples were allowed to react at 60 C for 6 h. Sample pressure was adjusted to 1 atm by allowing the sample to equilibrate with a helium-filled syringe connected through the vial septum. Evacuation, acid addition, and heated reaction were performed by the automated gas handling system developed for the CFAMS GIS carbonate system [@Roberts2013].

@@

Samples were measured on a modified 3MV Tandetron AMS system [@Longworth2015]. The displacement helium was allowed to flow constantly during analysis, and the sample needle was inserted into a helium-filled vial between samples. Measurement of a sample proceeded as follows: a clean sputter target cathode was rotated into position and the target was pre-cleaned by sputtering for 5 minutes or until cathode ^12^C^-^ current fell below 100 nA. The delivery needle was then transferred to the sample vial, which started sample delivery to the split and source. Currents were allowed to stabilize for 30 s , at which point the data acquisition was started. Data was acquired in 10 blocks of 90 s each, for 15 min total acquisition time. Normalizing standards, blanks, and secondary standards were measured identically to unknowns. Total time for a measurement was 20-25 min including target pre-cleaning and sample changes.

@@

After acquisition, data reduction, normalization and blank correction result in a final fraction modern with error. First, each 90 s block was corrected for fractionation using the AMS ^13/12^C measurement. Blocks that are more than 1.5 times the interquartile range (IQR) below the first quartile of all measurements of a sample or more than 1.5 IQR above the third quartile are considered outliers and were dropped at this point. Remaining runs of a sample were averaged and sample uncertainty was taken as the greater of counting error or the standard error of the measurements. The sample ratio was then normalized using the mean of gas standards measured during the run. The measurement error of the standards was taken as the greater of the standard error of all standards or the errors of the standards added in quadrature.

@@

# Results

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO~2~ connected through a regulator and glass capillary to the ion source to measure source performance over a range of CO~2~ flows. Best currents for the source was 10 $\mu$A ^12^C^-^, or roughly 1/10 of graphite target current. Peak current was produced at a CO~2~ flow rate of 7.5 $\mu$l min^-1^ CO~2~, but the source produced near optimal currents with flows from 7.5-25 $\mu$l min^-1^ CO~2~. This is comparable with other hybrid sources [@Xu2007], although improved target design and continued optimization of source parameters may improve performance [@Salazar2013; @Fahrni2013]. Typical currents for this study were between 4-9 $\mu$A and average current for all samples was `r curmean` ± `r cursd` $\mu$A. Several studies have shown best currents at lower CO~2~ flow rates [@Xu2007; @Ruff2010].

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig2}Performance of the NOSAMS MCSNICS hybrid gas ion source. (a) Measured $^{12}C^-$ ion current during displacement of a pure $CO_2$ vial sample with helium. (b) Modeled flows of helium and $CO_2$ to the ion source during the same dilution experiment. (c) Measured ion current for modeled $CO_2$ flow. The stability of the source over a range of flow produces stable current for up to 40 min of measurement time.", echo = FALSE}
knitr::include_graphics("../figures/fig2_current_flow.pdf")
```

@@

```{r fig3, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig3}Assessment of blanks. (a) Current dependent blank model and data. $F^{14}C$ and current of measurements from dilution tests of live and dead gas are used to generate linear fits which should cross at the $F^{14}C$ and current of the constant current blank. Direct determination of $F^{14}C$ and current of the blank are also shown. (b) Fraction modern of large, radiocarbon dead samples compared to sample current. Horizontal and vertical bars represent measurement uncertainty in current and ratio, respectively.", echo = FALSE}
knitr::include_graphics("../figures/fig3_blanks.pdf")
```

@@

```{r fig4, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig4}Agreement of standards with graphite consensus values. Difference of HGIS measurements from consensus for individual samples are plotted against their expected value. Expected values are spread to avoid overlapping points. Error bars represent one sigma sample uncertainty. The black line is the mean difference from expected value for all measurements, and the blue lines are one standard deviation of all measurements.", echo = FALSE}
knitr::include_graphics("../figures/fig4_consensus_samples.pdf")
```

@@

```{r fig5, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig5}Comparison of Haiti core samples measured as graphite and via HGIS. (a) Difference between HGIS and graphite measurements of the same materials. Difference between HGIS and graphite measurements is shown on the vertical axis and plotted against the mean of the two measurements. Samples are offset on the horizontal axis for clarity. Error bars are measurement error for each sample. Graphite measurements are included to allow comparison of error estimates. The bottom row of plots show a comparison of age-depth models produced using HGIS (b) and graphite measurements (c).", echo = FALSE}
knitr::include_graphics("../figures/fig5_core_carbonates.pdf")
```

@@

# Discussion

>>>>

Radiocarbon Accelerator Mass Spectrometry (AMS) is a sensitive method for determining the age, provenance, and composition of carbon-containing materials. A primary disadvantage of the method is its cost, which often limits the number of samples analyzed. This forces researchers to carefully choose samples to be run and limits repeat measurements for error determination. One driver of the cost of AMS analysis is the time and effort required for sample preparation, including reduction of the sample to graphite. Analysis of CO~2~ directly via a gas-accepting AMS ion source eliminates the cost and effort of graphitization and returns results more rapidly. The NOSAMS USAMS system at the National Oceanographic AMS lab (NOSAMS) is based on a 3MV Tandetron accelerator [@Longworth2015] combined with a National Electrostatics Corporation (NEC) 134-sample gas-capable MCSNICS source [@Xu2007] . We present here a system for semi-automated analysis of carbonate minerals by acid hydrolysis in vials and introduction to the ion source in a flow of helium via an open split.

@@

```{r fig1, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig1}Schematic of the open split inlet. Two valco 2-way valves provide the ability to admit helium and live or dead tank standards. The double needle is housed in a VICI cross and allows displacement helium to flow into the sample vial and the sample-helium mixture to flow to the open split, where a capillary delivers the mixture to the ion source.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig1_HGIS_open_split_drawing.pdf"))
```

@@

Carbonate samples were weighed into 7 ml septa-sealed Vacutainer vials (Labco 768W). Sealed vials were evacuated, then 1 ml 85% phosphoric acid was added. The samples were allowed to react at 60 C for 6 h. Sample pressure was adjusted to 1 atm by allowing the sample to equilibrate with a helium-filled syringe connected through the vial septum. Evacuation, acid addition, and heated reaction were performed by the automated gas handling system developed by Roberts et al. [@Roberts2013].

@@

Samples were measured on a modified 3MV Tandetron AMS system [@Longworth2015]. The displacement helium was allowed to flow constantly during analysis, and the sample needle was inserted into a helium-filled vial between samples. Measurement of a sample proceeded as follows: a clean sputter target cathode was rotated into position and the target was pre-cleaned by sputtering for 5 minutes or until cathode ^12^C^-^ current fell below 100 nA. The delivery needle was then transferred to the sample vial, which started sample delivery to the split and source. Currents were allowed to stabilize for 30 s, at which point the data acquisition was started. Data was acquired in 10 blocks of 90 s each, for 15 min total acquisition time. Normalizing standards, blanks, and secondary standards were measured identically to unknowns. Total time for a measurement was 20-25 min including target pre-cleaning and sample changes.

@@

After acquisition, data reduction, normalization and blank correction result in a final fraction modern with error. First, each 90 s block was corrected for fractionation using the AMS ^13/12^C measurement. Blocks that are more than 1.5 times the interquartile range (IQR) below the first quartile of all measurements of a sample or more than 1.5 IQR above the third quartile are considered outliers and were not further used in the analysis. Remaining runs of a sample were averaged and sample uncertainty was taken as the greater of counting error or the standard error of the measurements. The sample ratio was then normalized using the mean of gas standards measured during the run. The measurement error of the standards was taken as the greater of the standard error of all standards or the errors of the standards added in quadrature.

@@

# Results and Discussion

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO~2~ connected through a regulator and glass capillary to the ion source to measure source performance over a range of CO~2~ flows. Best currents for the source was 10 $\mu$A ^12^C^-^, or roughly 10% of graphite target current. Peak current was produced at a CO~2~ flow rate of 7.5 $\mu$l min^-1^ CO~2~, but the source produced near optimal currents with flows from 7.5-25 $\mu$l min^-1^ CO~2~. This is comparable with other hybrid sources [@Xu2007], although improved target design and continued optimization of source parameters may improve performance [@Salazar2013; @Fahrni2013]. Typical currents for this study were between 4-9 $\mu$A and average current for all samples was `r curmean` ± `r cursd` $\mu$A. Several studies have shown best currents at lower CO~2~ flow rates [@Xu2007; @Ruff2010].

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig2}Performance of the NOSAMS MCSNICS hybrid gas ion source. (a) Measured $^{12}C^-$ ion current during displacement of a pure $CO_2$ vial sample with helium. (b) Modeled flows of helium and $CO_2$ to the ion source during the same dilution experiment. (c) Measured ion current for modeled $CO_2$ flow. The stability of the source over a range of flow produces stable current for up to 40 min of measurement time.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig2_current_flow.pdf"))
```

@@

```{r fig3, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig3}Assessment of blanks. (a) Current dependent blank model and data. $F^{14}C$ and current of measurements from dilution tests of live and dead gas are used to generate linear fits which should cross at the $F^{14}C$ and current of the constant current blank. Direct determination of $F^{14}C$ and current of the blank are also shown. (b) Fraction modern of large, radiocarbon dead samples compared to sample current. Horizontal and vertical bars represent measurement uncertainty in current and ratio, respectively.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig3_blanks.pdf"))
```

@@

```{r fig4, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig4}Agreement of standards with graphite consensus values. Difference of HGIS measurements from consensus for individual samples are plotted against their expected value. Expected values are spread to avoid overlapping points. Error bars represent one sigma sample uncertainty. The black line is the mean difference from expected value for all measurements, and the blue lines are one standard deviation of all measurements.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig4_consensus_samples.pdf"))
```

@@

```{r fig5, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig5}Comparison of Haiti core samples measured as graphite and via HGIS. (a) Difference between HGIS and graphite measurements of the same materials. Difference between HGIS and graphite measurements is shown on the vertical axis and plotted against the mean of the two measurements. Samples are offset on the horizontal axis for clarity. Error bars are measurement error for each sample. Graphite measurements are included to allow comparison of error estimates. The bottom row of plots show a comparison of age-depth models produced using HGIS (b) and graphite measurements (c).", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig5_core_carbonates.pdf"))
```

@@

# Conclusions
